---
title: "R Notebook"
output: html_notebook
Prepared by: kimnewzealand
Date: 23 November 2017
---

## Synopsis

The objective of this analysis is to use a Billboard lyrics dataset and perform sentiment analysis across the decades using sentiment plots.

### 

 

* * *

## Setup

```{r setup, results='asis', include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE)
```

**LOAD R PACKAGES**

```{r warnings=FALSE,include=FALSE}
library(tidyverse) # this loads the related tidyverse packages including dplyr, ggplot2 and readr
library(tidytext)
library(radarchart)
library(htmlwidgets)
library(htmltools)
```

**LOAD DATA**

Load the dataset already prepared in this [datacamp link](https://assets.datacamp.com/production/course_3424/datasets/song_lyrics.rda) originally sourced by Kaylin Walker. 

This dataset contains over 5000 songs, from 1965 to 2015. 

song_lyrics.Rda has the following variables:

- rank, the rank a song achieved on the Billboard Year-End Hot 100
- song, the song's title
- artist, the artist who recorded the song
- year, the year the song reached the given rank on the Billboard chart
- lyrics, the lyrics of the song.

```{r load}
url="https://assets.datacamp.com/production/course_3424/datasets/song_lyrics.rda"
download.file(url,"song_lyrics.Rda")
load("song_lyrics.Rda")
```

## Clean Data

First we will convert the song_lyrics into a tibble, then into tidy dataset. We will use the song as the variable to group by and then extract the words from the lyrics as the output column from the input column lyrics using unnest_tokens function from [tidytext](https://cran.r-project.org/web/packages/tidytext/index.html) R package.

```{r cleandata}
# Convert to tibble
song_lyrics <- as_tibble(song_lyrics[c(2:5)],)
# Tidy dataset
tidy_lyrics <-      song_lyrics %>%
# Group by the songs and years
  group_by(song,year) %>%
# Define a new column linenumber
  mutate(linenumber=row_number()) %>%
# Transform the non-tidy text data to tidy text data. Default "words" token.
  unnest_tokens(word, lyrics) %>%
  ungroup()
```

Perform sentiment Analysis using the bing and nrc lexicons.

```{r sentimentbing}
# Sentiment analysis using the bing lexicon
lyric_sentiment_bing <-  tidy_lyrics %>%
  # Implement sentiment analysis with the "bing" lexicon
  inner_join(get_sentiments("bing")) 

lyric_tidy_sentiment <- lyric_sentiment_bing %>% 
  count(word, sentiment) %>%
  spread(sentiment, n, fill = 0) %>%
  mutate(polarity = positive -negative)

# Let's filter by polarity score above 500 and remove offensive words
lyrics_tidy_small <- lyric_tidy_sentiment %>% 
  filter(abs(polarity) >= 500) %>% 
  filter(word!="fuck") %>% 
  filter(word!="shit") %>% 
  filter(word!="bitch")

# Add polarity descriptions
lyrics_tidy_pol <- lyrics_tidy_small %>% 
  mutate(Polarity = ifelse(polarity > 0, "positive", "negative"))

```

```{r sentimentnrc}
# Sentiment analysis using the nrc lexicon
lyric_sentiment_nrc <-  tidy_lyrics %>%
  # Implement sentiment analysis with the "n" lexicon
  inner_join(get_sentiments("nrc")) 
# Group years into decades
lyric_tidy_sentiment_count_nrc <- lyric_sentiment_nrc %>% 
      mutate(Decade = ifelse(year<1970,"1960s",ifelse(year>2010,"2010s",ifelse(year>1999,"2000s",ifelse(year>1989,"1990s",ifelse(year>1979,"1980s",ifelse(year>1969,"1970s","no"))))))) %>%
  group_by(sentiment,Decade) %>%
      count(sentiment) %>%
      spread(Decade,n)
```

## Exploratory Data Analysis

**Word frequency barplot**

We use the [tidytext](https://cran.r-project.org/web/packages/tidytext/index.html) R package to perform sentiment analysis using the bing lexicon, which categorises words in a binary fashion into positive and negatives.

Then, plot a word frequency bar plot of the word counts on the positive and negative sentiment.

```{r barplot}
# Plot a barplot of the Sentiment Word Frequency, with the coordinates flipped
freqplot <- lyrics_tidy_pol %>% ggplot(aes(reorder(word, polarity), polarity, fill = Polarity)
) +  geom_bar(stat = "identity") + 
  ggtitle("Billboard Song Lyrics: Sentiment Word Frequency") + 
      xlab("Words") + 
      ylab("Count") + 
       coord_flip() + 
      theme(axis.text=element_text(size=7))
freqplot# Save to disk using ggplot2 ggsave function
ggsave("Billboardfreqplot.png")
```

**Radar Chart**

Next we will take a look at a Radar Chart of the sentiments by decades.

Next, we plot radar charts using the [radarchart](https://cran.r-project.org/web/packages/radarchart/index.html) R package. Radarchart creates an html interactive chart, embedded below.

We compare the sentiment between the decades using the ten emotion category NRC lexicon.

```{r radarchart}
# JavaScript radar chart default package settings
decades<- chartJSRadar(lyric_tidy_sentiment_count_nrc)
#plot the html chart
decades
# save the html as a widget
saveWidget(decades,"BillboardDecades.html")
# Create a colour matrix to customise the colours in the chart
cols <- col2rgb(c("darkblue", "darkgreen", "red"))
# plot the Radar chart using a narrow format, splitting the decades into two groups
decades1<- chartJSRadar(lyric_tidy_sentiment_count_nrc[,1:4],maxScale = 15000, scaleLineWidth=10,addDots = TRUE, colMatrix = cols) #, colMatrix = cols)
decades1
saveWidget(decades1,"BillboardDecades1.html")
decades2<- chartJSRadar(lyric_tidy_sentiment_count_nrc[,c(1,5:7)],maxScale = 25000, scaleLineWidth=10,addDots = TRUE, colMatrix = cols) #, colMatrix = cols)
decades2
saveWidget(decades2,"BillboardDecades2.html")
```

