---
title: "World Bank data with tmap"
output: html_notebook
Date: 30 November 2017
AUthor: kimnewzealand
---

## SYNOPSIS

We will be sourcing World Development Indicators from World Bank [open data](https://data.worldbank.org/) to compare communication technologies between countries in the developed, emerging and developing economies.

With this data we will plot interactive thematic maps.

### 

 

* * *

## Setup

**LOAD R PACKAGES**

```{r libraries,warnings=FALSE,include=FALSE}
library(WDI) 
library(wbstats)
library(data.table)
library(RColorBrewer)
library(tmap)
library(tmaptools)
library(leaflet)
library(leaflet.extras)
library(htmlwidgets)

```
  
### 

 

* * *

## Data

**LOAD WORLD BANK DATA**

First we will load our World Development Indicators (World Bank) data from the [WDI](https://cran.r-project.org/web/packages/WDI/index.html) R package. The wb function from [wbstats](https://cran.r-project.org/web/packages/wbstats/index.html) R package downloads the requested information using the World Bank API.

```{r load WDI}
# World Development Indicators (sourced frWorld Bank)
wpp<-WDIsearch("per 100 people")
head(wpp)
# Note as per https://data.worldbank.org/indicator/IT.NET.USER.P2 The indicator IT.NET.USER.P2 does not exist or is no longer available
# Note as per https://data.worldbank.org/indicator/IT.CMP.PCMP.P2 The indicator IT.CMP.PCMP.P2 does not exist or is no longer available
# Source the indicators and the corresponding values into a dataset which is appended by rows
commdata <- data.table(
  wb(indicator = c("IT.CEL.SETS.P2", # Mobile cellular subscriptions (per 100 people)
                   "IT.MLT.MAIN.P2", # Telephone lines (per 100 people)
                   "IT.NET.BBND.P2"), # Fixed broadband subscriptions (per 100 people)
     mrv = 1) # mrv is the number of Most Recent Values to return.
)  
# Take a look at the data summary
str(commdata)
# check that the data has loaded for each indicator
table(commdata$indicatorID)
```


**LOAD WORLD MAP DATA**

Next we will load the "World" data which is part of the [tmap](https://cran.r-project.org/web/packages/tmap/index.html) R package.

```{r world dataset}
# Load World dataset
data("World")
# Take a look at the class and structure of World
str(World,max.level=2)
# Take a look at the variables in the data slot
str(World@data,max.level=2)
```
World is a class `r summary(World)[2]`.

**CLEAN WDI DATA**

The World Data Indicator dataset includes a country variable but this is at a different level and aggregation eg "Arab World", "High Income" to the "World" dataset so we need to remap the country variables with the wbcountries function from the wbstats R package.

```{r clean WDI}
# Download country mappings
countries <- data.table(wbcountries())
# Set keys to join the data sets
setkey(commdata, iso2c)
setkey(countries, iso2c)
# Add regions to the data set, but remove aggregates
commdata<- countries[commdata][ ! region %in% "Aggregates"]
commdata <- commdata %>% 
      dplyr::select(country,indicatorval=value,indicatorID,indicator) # rename the value to indicatorval
# Check for missing values in our dataset by column
apply(is.na(commdata),2,sum)
```


**MERGE DATASETS**

Next we will append the commdata dataframe to the the World shape object using append_data function from the [tmaptools](https://cran.r-project.org/web/packages/tmaptools/index.html) R package. We will create different objects for each indicator so that there is a single country in each object to map. Assume that the countries that are not mapped do not have WDI data available.

```{r append}
# Match the variables name in the World object and country in the commdata for mobile indicator
mobile <- append_data(World,commdata[commdata$indicatorID== "IT.CEL.SETS.P2",],key.shp="name",key.data="country") 
# Match the telephone indicator
telephone <- append_data(World,commdata[commdata$indicatorID== "IT.MLT.MAIN.P2",],key.shp="name",key.data="country")
# Match the broadband indicator
bband <- append_data(World,commdata[commdata$indicatorID== "IT.NET.BBND.P2",],key.shp="name",key.data="country")
```


### 

 

* * *

## Visualisations

**EXPLORATORY DATA ANALYSIS**

We will initially plot World using the [tmap](https://cran.r-project.org/web/packages/tmap/index.html) R Package plotting functions. This map will colour the economy variable provided in the World dataset.

```{r first tmap}
# Plot a static map
tmap_mode("plot")

worldecon <- tm_shape(World) + # Specifies the shape object
      tm_borders() + # Add layer to the map with borders of each country polygon
      tm_fill(col="economy") + # Add a layer with the fill mapped to the economy variable
      tm_compass() + # Add compass icon
      tmap_style("cobalt") + # Set the tmap style 
      tm_layout(legend.position = c("left","bottom")) # Reset the legend to bottom left
# Print map
worldecon
# Change the colour palatte with 7 steps on the RColorBrewer "YlOrBr" palette
colpal <- rev(brewer.pal(n=7,"YlOrBr"))
# Replot with a customised sequential palette
worldecon2 <- tm_shape(World) + # Specifies the shape object
      tm_borders() + # Add layer to the map with borders of each country polygon
      tm_fill(col="economy",palette = colpal) + # Add a layer with the fill mapped to the economy variable
      tm_compass() + # Add compass icon
      tmap_style("cobalt") + # Set the tmap style 
      tm_layout(legend.position = c("left","bottom")) # Reset the legend to bottom left
worldecon2
# Save the worldecon map to disk
save_tmap(tm = worldecon2,"WorldEconomies2.png")
```

**DATA VISUALISATION**

```{r}
# Set mode to plot a static map with the default colour palette 
tmap_mode("plot")
# Plot the MOBILE indicator and use the default colour palette
mobilemap<- tm_shape(mobile) + # Specifies the shape object
      tm_borders() + # Add layer to the plot with borders of each country polygon
      tm_fill(col="indicatorval", title=("Mobile (per 100 people)"),
              legend.hist=FALSE) + # Add a layer with the fill mapped to the indicator value variable
      tmap_style("cobalt") +
      tm_layout(legend.position = c("left","bottom")) # Reset the legend to bottom left
mobilemap
# Change tmap to interactive
tmap_mode("view")
mobilemapinteract <- tmap_leaflet(mobilemap)
# Reset the centering and zoom with leaflet setView
mobilemapinteract<- mobilemapinteract %>% setView(0,0,zoom = 1)
saveWidget(mobilemapinteract,"mobilemap.html")
#
# Set mode to plot a static map 
tmap_mode("plot")
# Change the colour palatte with 6 steps on the RColorBrewer "PuBu" palette
colpal <- brewer.pal(n=6,"PuBu")
# Plot the TELEPHONE indicator and use the customised colour palette
telmap<-tm_shape(telephone) + # Specifies the shape object
      tm_borders() + # Add layer to the plot with borders of each country polygon
      tm_fill(col="indicatorval",palette = colpal, title=("Telephone (per 100 people)"),legend.hist=FALSE) + # Add a layer with the fill mapped to the indicator value variable
      tmap_style("cobalt") +
      tm_layout(legend.position = c("left","bottom")) # Reset the legend to bottom left
telmap
# Change tmap to interactive
tmap_mode("view")
# Saves as a leaflet map
telmapinteract <- tmap_leaflet(telmap) 
# Reset the centering and zoom with leaflet setView
telmapinteract<- telmapinteract %>% setView(0,0,zoom = 1) 
saveWidget(telmapinteract,"telmap.html")
#
# Set mode to plot a static map 
tmap_mode("plot")
# Change the colour palatte with 6 steps on the RColorBrewer "RdPu" palette
colpal <- brewer.pal(n=6,"RdPu")
# Plot the BROADBAND indicator and use the customised colour palette
bbandmap <- tm_shape(bband) + # Specifies the shape object
      tm_borders() + # Add layer to the plot with borders of each country polygon
      tm_fill(col="indicatorval",palette = colpal, title=("Broadband (per 100 people)")) + # Add a layer with the fill mapped to the continent variable
      tmap_style("cobalt") +
      tm_layout(legend.position = c("left","bottom")) # Reset the legend to bottom left
bbandmap
# Change tmap to interactive
tmap_mode("view")
# Saves as a leaflet map
bbandinteract <- tmap_leaflet(bbandmap) 
# Reset the centering and zoom with leaflet setView
bbandinteract <- bbandinteract %>% setView(0,0,zoom = 1) 
saveWidget(bbandinteract,"bband.html")
bbandinteract

```

